Retrofitting Effect Handlers onto OCaml

### 2021-03-31 retro-concurrency.tex
Update for camera ready

\newcommand{\caseval}[2]{\kw{return}~#1 \mapsto #2}
\newcommand{\caseexn}[3]{\kw{exception} ~#1 ~#2 \mapsto #3}
\newcommand{\caseeff}[4]{\kw{effect} ~#1 ~#2 ~#3 \mapsto #4}

@@

\begin{figure*}
\begin{minipage}{0.30\linewidth}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[numbers=left]
external ocaml_to_c
  : unit -> int = "ocaml_to_c"
exception E1
exception E2
let c_to_ocaml () = raise E1
let _ = Callback.register
    "c_to_ocaml" c_to_ocaml
let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c ()
    | with E2 -> 0
  | with E1 -> 42;;
let _ = assert (omain () = 42)
    \end{lstlisting}
    \vspace{-2mm}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[language=c,numbers=left]

#include <caml/mlvalues.h>
#include <caml/callback.h>

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push k runq in
  let rec run_next () =
    match Queue.pop runq with
    | k -> continue k ()
    | exception Queue.Empty -> ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
  in
  spawn main
\end{lstlisting}

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (In_line ic) k ->
      pending_reads := (ic,k)::!pending_reads; run_next ()
  in
  spawn main
\end{lstlisting}

@@

\section{Semantics}
\label{sec:semantics}

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect safety~\cite{Eff,
Hillerstrom20, Leijen14, Effekt, Frank, Biernacki19}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases; programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return field |retc| of type |'a -> 'b|. The effect field
|effc| handles effects of type |'c eff| with |('c,'b) continuation| and returns
a value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperform is implemented as a
primitive to avoid executing code on the resumption path.

@@

@@

		\begin{array}{cc}

@@

								{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &

@@

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
	\vspace{-4mm}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\vspace{-4mm}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_size}
\end{figure*}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{retro-concurrency}

\if{0}
\clearpage

\section*{Supplementary Material}
\setcounter{figure}{0}

\subsection*{Dynamic Semantics}

\begin{definition}[Initial configuration]
	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
\end{definition}

Intuitively, the initial environment is empty and the initial stack is an empty
C stack.

\begin{definition}[Final value configuration]
	Given a configuration $\config$, we say that the configuration is the final
	value configuration (written $\textsf{finalval}(\config)$) if ~$\config =
	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
	$\env$.
\end{definition}

\begin{definition}[Final exception configuration]
	Given a configuration $\config$, we say that the configuration is the final
	exception configuration (written $\textsf{finalexn}(\config)$) if ~$\config
	= \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl}{\bullet}}$~ for some
	value $v$, label $l$, frame list $\fl$ and environment $\env$.
\end{definition}

Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
the transitive reflexive closure of the top-level reduction relation $\step$.

\subsubsection*{Example 1}

Let

\begin{smathpar}
	\begin{array}{rcll}
		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
			&   &            & \mid \caseval{v}{v}})~0
	\end{array}
\end{smathpar}

then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
42$.

\begin{proof}
\end{proof}

\subsection*{Experimental support for multi-shot continuations}

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming (for example, see Uber's Pyro language~\cite{Pyro}). Multicore
OCaml has an experimental feature for explicitly cloning continuations, which
makes a copy of each of its constituent fibers.

Multi-shot continuations are difficult to
reason about in the presence of linear resources such as file descriptors~\cite{de-vilhena-pottier-21}.
Recall that OCaml programs are written with the expectation that function calls
return \emph{exactly-once} either normally or exceptionally. With multi-shot
continuations, they may return \emph{more than once}, leading to errors such as
double-freeing malloced memory. Since our goal is to retrofit effect handlers
onto OCaml, we relegate continuation cloning to the OCaml |Obj| module, which
contains unsafe features.

In addition, multi-shot continuations currently are not safe under compiler
optimisations. For example consider the following program:

\begin{minipage}{\linewidth}
\begin{lstlisting}
effect E : unit
let foo () = perform E
let bar f =
  let r = ref 0 in
	f ();
  r := !r + 1;
  !r
;;
match bar foo with
| v -> v
| effect E k ->
    let k' = Obj.clone_continuation k in
    let r1 = continue k () in
    let r2 = continue k' () in
    Printf.printf "%d %d\n" r1 r2;
		0
\end{lstlisting}
\end{minipage}

This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
the presence of multi-shot continuations, compiler optimisations need to be
made aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.
\fi

>>>>

\newcommand{\caseval}[2]{\kw{return}~#1 \mapsto #2}
\newcommand{\caseexn}[3]{\kw{exception} ~#1 ~#2 \allowbreak \mapsto #3}
\newcommand{\caseeff}[4]{\kw{effect} ~#1 ~#2 ~#3 \mapsto #4}

@@

\begin{figure*}
\begin{minipage}{0.30\linewidth}
  \begin{minipage}{\linewidth}
		\begin{lstlisting}[xleftmargin=0.05\linewidth]
external ocaml_to_c
  : unit -> int = "ocaml_to_c"
exception E1
exception E2
let c_to_ocaml () = raise E1
let _ = Callback.register
    "c_to_ocaml" c_to_ocaml
let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c ()
    | with E2 -> 0
  | with E1 -> 42;;
let _ = assert (omain () = 42)
    \end{lstlisting}
    \vspace{-2mm}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
		\begin{lstlisting}[language=c, xleftmargin=0.05\linewidth]
#include <caml/mlvalues.h>
#include <caml/callback.h>

@@

\begin{lstlisting}[numbers=left, xleftmargin=0.025\textwidth]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push k runq in
  let rec run_next () =
    match Queue.pop runq with
    | k -> continue k ()
    | exception Queue.Empty -> ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
  in
  spawn main
\end{lstlisting}

@@

\begin{lstlisting}[numbers=left, xleftmargin=.025\textwidth]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (s,k) ->
         Queue.push (fun () -> continue k s) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (In_line ic) k ->
      pending_reads := (ic,k)::!pending_reads; run_next ()
  in
  spawn main
\end{lstlisting}

@@

\vspace{-1mm}
\section{Semantics}
\label{sec:semantics}

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety. Programs without matching effect handlers are well-typed
Multicore OCaml programs. As a result, our static semantics is simpler than
languages that ensure effect safety~\cite{Eff, Hillerstrom20, Leijen14, Effekt,
Frank, Biernacki19}. This is important for backwards compatibility as our goal
is to retrofit effect handlers to a language with large legacy codebases;
programs that do not use effects remain well-typed, and those that do compose
well with those that don't.

@@

For effect handling, we extend OCaml's |match ... with| syntax with effect
patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return field |retc| of type |'a -> 'b|. The effect field
|effc| handles effects of type |'c eff| with |('c,'b) continuation| and returns
a value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperform is implemented as a
primitive to avoid executing code on the resumption path.

@@

\begin{minipage}{0.25\linewidth}

@@

		\begin{array}{c}

@@

								{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\

@@

\begin{minipage}{0.69\linewidth}

@@

%\vspace{-1mm}
\subsection{External calls and callbacks}

@@

%\vspace{-1mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\begin{figure*}
	\includegraphics[scale=0.34]{sandmark-notebook/sandmark_time}
	\vspace{-4mm}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.34]{sandmark-notebook/sandmark_codesize}
	\vspace{-4mm}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_size}
\end{figure*}

@@

\bibliographystyle{ACM-Reference-Format}
\balance
\bibliography{retro-concurrency}
